---
title: "Locating Chromothripsis Events with ShatterSeek"
author: "Kelsey Luu"
date: "3/15/2021"
output: html_document
---

```{r setup, include=FALSE}
library(gridExtra)
library(cowplot)
library(tidyverse)
library(ggplot2)
library(gplots)
library(scales)
library(data.table)
```


### Pre-process data
```{r, eval=F}
# read in SV and CN dfs:
SV_file <- "inputs/WCDT101_svaba-titan.bedpe"
CN_file <- "inputs/CRPC10X42-WCDT101_titanIchor.seg.joined.txt"
primary <- FALSE  # TRUE for primary tumors
SV_df <- read.table(SV_file, header = T, sep = '\t')
CN_df <- read.table(CN_file, header = T, sep = '\t') %>% mutate(len = End - Start)

all_samples <- as.character(CN_df$Sample) %>% unique()

if (exists("SV_df") & exists("CN_df")) {
  if (primary) {
    # set col names (only for primary tumors)
    names(SV_df)[c(1:6, 9:11)] <- c("chrom1", "start1", "end1",
                                    "chrom2", "start2", "end2",
                                    "strand1", "strand2", "Sample")
    names(CN_df) <- c("Sample", "Chromosome", "Start", "End", "Corrected_Copy_Number", "Length")
  }
  
  # fix chromosome numbers
  SV_df$chrom1 <- gsub("[^(0-9) | X]", "",  SV_df$chrom1) 
  SV_df$chrom2 <- gsub("[^(0-9) | X]", "",  SV_df$chrom2)
  CN_df$Chromosome <- gsub("[^(0-9) | X]", "",  CN_df$Chromosome)
  chrs <- c(1:22, "X")  # only include chr1~22 and X
  SV_df <- SV_df %>% filter(chrom1 %in% chrs, chrom2 %in% chrs)
  CN_df <- CN_df %>% filter(Chromosome %in% chrs, !is.na(Corrected_Copy_Number))
  
  #  add SV class to SV_df
  DUP_id <- which(SV_df[,"strand1"] == "-" & SV_df[,"strand2"] == "+")
  DEL_id <- which(SV_df[,"strand1"] == "+" & SV_df[,"strand2"] == "-")
  h2hINV_id <- which(SV_df[,"strand1"] == "+" & SV_df[,"strand2"] == "+")
  t2tINV_id <- which(SV_df[,"strand1"] == "-" & SV_df[,"strand2"] == "-")
  SV_df[DUP_id,"SVtype"] <- "DUP"
  SV_df[DEL_id,"SVtype"] <-"DEL"
  SV_df[h2hINV_id,"SVtype"] <- "h2hINV"
  SV_df[t2tINV_id,"SVtype"] <- "t2tINV"
}

# run ../scripts/join_CNsegments.R on all.seg.smooth.txt and read in output
CN_df <- fread(sprintf('inputs/%s.joined.txt', gsub(".txt", "", basename(CN_file))))
```


###  Run shatterseek 
```{r, eval=F}
chromoth_list <- list()
for (sample in all_samples) {
  # subset sample
  sample_SV <- SV_df %>% filter(Sample == sample) 
  sample_CN <- CN_df %>% filter(Sample == sample)
  
  # load SV data into SV object
  SV_data <- SVs(chrom1=as.character(sample_SV$chrom1),
                 pos1=as.numeric(sample_SV$start1),
                 chrom2=as.character(sample_SV$chrom2),
                 pos2=as.numeric(sample_SV$end2),
                 strand1=as.character(sample_SV$strand1),
                 strand2=as.character(sample_SV$strand2),
                 SVtype=as.character(sample_SV$SVtype))
  
  # load CNV data into CSVsegs object
  CN_data <- CNVsegs(chrom=as.character(sample_CN$Chromosome),
                     start=sample_CN$Start,
                     end=sample_CN$End,
                     total_cn=sample_CN$Corrected_Copy_Number)
  
  # find chromothripsis
  start_time <- Sys.time()
  chromothripsis <- shatterseek(SV.sample=SV_data, seg.sample=CN_data)
  chromothripsis@chromSummary <- chromothripsis@chromSummary %>% mutate(sample=sample)
  end_time <- Sys.time()
  print(paste0("Running time (s): ", round(end_time - start_time, digits=2)))
  
  # save chromoth object
  chromoth_list <- c(chromoth_list, chromothripsis)
}

# save output to rds
# saveRDS(chromoth_list, file = "outputs/shatterseek_chromoth_out.seg.joined.RDS")
```


### Save shatterseek plots to pdf
```{r, eval=F}
fig_dir <- "./plots/ShatterSeek_plots/all_plots/"

chrom_plot <- function(chromo_obj, chr) {
  plots = plot_chromothripsis(ShatterSeek_output = chromo_obj,chr = chr)
  plot_out = arrangeGrob(plots[[1]],
                         plots[[2]],
                         plots[[3]],
                         plots[[4]],
                         nrow=4,ncol=1,heights=c(0.2,.4,.4,.4))
  return(plot_out)
}

for (chrom_obj in chromoth_list) {
  df <- chrom_obj@chromSummary[!is.na(chrom_obj@chromSummary$start),]
  sample_id <- df$sample[1]
  plotdir <- paste0(fig_dir, sample_id)
  dir.create(plotdir)
  for (chr in df$chrom) {
    figpath <- paste0(plotdir, "/", sample_id, "-chr", chr, ".pdf")
    pdf(figpath)
    plot(chrom_plot(chromo_obj = chrom_obj, chr = chr))
    dev.off()
  }
}
```


## Filtering criteria for WCDT
### Compile summary table with confidence classifications

For WCDT filtering, we adopted and refined confidence classification criteria from the ShatterSeek documentation: https://github.com/parklab/ShatterSeek/blob/master/tutorial.pdf 
```{r, eval=F}
summary_df_list <- lapply(chromoth_list,
                          function(x) x@chromSummary %>% filter(!is.na(start)))
alpha <- 0.05  # significance threshold

full_summary <- bind_rows(summary_df_list) %>%
  mutate(confidence =
    case_when(
      clusterSize_including_TRA - number_TRA >= 10 &
        max_number_oscillating_CN_segments_2_states >= 10 &
        (chr_breakpoint_enrichment <= alpha | pval_exp_cluster <= alpha)
      ~ "high",
      clusterSize_including_TRA - number_TRA >= 3 &
        number_TRA >= 4 &
        max_number_oscillating_CN_segments_2_states >= 7 &
        (chr_breakpoint_enrichment <= alpha | pval_exp_cluster <= alpha)
      ~ "medium",
      clusterSize_including_TRA - number_TRA >= 6 &
        between(max_number_oscillating_CN_segments_2_states, 4, 6) &
        (chr_breakpoint_enrichment <= alpha  | pval_exp_cluster <= alpha)
      ~ "low"
    )
  ) %>% mutate(confidence = factor(confidence, levels = c("high", "medium", "low"))) %>%
  mutate(chrom = factor(chrom, levels = c(1:22, "X"))) 


# write all calls to summary table:
# full_summary %>% write.csv("outputs/prefilter_shatterseek_summary_table.seg.joined.csv", row.names = F)

# write putative real calls to summary table: 
# write.csv(full_summary %>% filter(!is.na(confidence)), "outputs/filtered_shatterseek_summary_table.seg.joined.csv", row.names = F)
```


## Filtering criteria for primary tumors & 10X
### Compile summary table with confidence classifications

For primary tumors and 10X, the above criteria yielded too many false positives. We suspected this was due to oversensitivity to CN oscillations. We instead filter these calls based on a weighted score that is primarily determined by the number of SVs in a cluster, with less weight given to CN oscillations.
```{r, eval=F}
summary_df_list <- lapply(chromoth_list,
                          function(x) x@chromSummary %>% filter(!is.na(start)))
alpha <- 0.05  # significance threshold

full_summary <- bind_rows(summary_df_list) %>%
    mutate(score =
             0.6 * as.integer(clusterSize_including_TRA - number_TRA >= 10) +
             0.2 * as.integer(max_number_oscillating_CN_segments_2_states >= 7 |
                        max_number_oscillating_CN_segments_3_states >= 14
                        ) +
             0.1 * as.integer(chr_breakpoint_enrichment < alpha) +
             0.1 * as.integer(pval_exp_chr < alpha)
    ) %>%
    mutate(confidence = case_when(
      score >= 0.8 ~ "high",
      score >= 0.7 ~ "medium",
      score >= 0.4 ~ "low"
    )
  ) %>% mutate(confidence = factor(confidence, levels = c("high", "medium", "low"))) %>%
  mutate(chrom = factor(chrom, levels = c(1:22, "X"))) 


# write all calls to summary table:
# full_summary %>% write.csv("outputs/prefilter_shatterseek_summary_table.seg.joined.csv", row.names = F)

# write putative real calls to summary table: 
# full_summary %>% filter(score >= 0.4) %>% write.csv("outputs/filtered_shatterseek_summary_table.seg.joined.csv", row.names = F)
```


### Visualize confidence distribution
```{r, eval=F}
full_summary %>% 
  filter(!is.na(confidence)) %>%
  ggplot() +
  geom_bar(aes(x = confidence)) +
  geom_text(stat="count", aes(x=confidence, label=..count..), vjust=-0.3) +
  labs(title = "Confidence Distribution")
# ggsave("plots/summary_plots/confidence_dist.pdf", height=5, width=6)
```

After this workflow, we manually curated the filtered ShatterSeek calls. 
